---
title: "2.13 hypothesis test"
author: "Jiahuan Li"
date: "2023-02-13"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# one tail test, alpha = 0.05
Z_crit = qnorm (p = 0.05, 0, 1, lower.tail = F)
# 1.64 that is the value of critical Z

# two tails test, alpha = 0.05
Z_crit2 = qnorm (p = 0.025, 0, 1, lower.tail = F)

# calculate p-value (with n = 30, don't use t test (n-1), use the z value instead)
set.seed(999)
unvacc = rpois(n = 30, lambda = 3)
mean = mean(unvacc)
sd = sd(unvacc)

# actual Z score get from real data, then compared to the Z distribution
Z = (mean - 0) / (sd / n)
pnorm(q = 2.9, mean = 0, sd = 1.69/sqrt(30), lower.tail = F)

# if Z_true (for population distribution)= 1.7. compare to different critical Z, it will be in the critical zone of the one tail test, but does not in the zone of two-tail test

# p value is the same and does not change according to hypotheses H1 or H2. But the critical Z will change according to different hypotheses conditions
```

one sample test: have one sample data two sample test: data from two samples and comparison

Assumption: independent; normally distributed/ two samples: approximately variance

pair test: measure the sample at different time (today/ after year obs of the same sample)

## t test

it is a one sample (compare to a value), one-sided (larger than) test

```{r}
## small sample size
sal = c(0.593, 0.142, 0.329, 0.691, 0.231, 0.793, 0.519, 0.392, 0.418)
t.test (x = sal, mu = 0.3, alternative = "greater")
```

conclusion: significant greater

t-value, df, mean should be reported in paper. only the p-value is not enough

##### what is the H0 in this situation? no greater than? not just unequal? link to the quiz error?

check on Wed

### t test 2-sample test

```{r}
cont = c(95, 94, 99, 96, 95, 96)
treat = c(101, 98, 100, 97, 99, 99)

# compare the cont (y) to the treat (x)? less means y < x?
## but it does not consist with the format of one-sample test?

t.test(x= treat, y = cont, alternative = "less")

# the below one hypo is correct
t.test(x= treat, y = cont, alternative = "greater")
# final answer: reject the h0, the hypothesis is true
```
