### Two sample tests

placebo (treatment & control): two sample, 1-sided

### Test hyphothesis hold or not: normality & equal variance

```{r}
cont = c(95, 94, 99, 96, 95, 96)

## test assumption: normality & equal variance
## H0 the data is normally distributed
## p-value: to reject the H0. 21% means cannot reject (p for wrong rejection)
shapiro.test(cont)
```

F test for equal variance: H0: their vars are equal

p = 0.67, is wrongly rejected H0, thus, the sds are equal

from the boxplot, look at the length/width of the two plots, they reflect the variance

### two sample test Result

one-side test: CI only has one number, the other side is infinity

t value 3.4 shown on the picture as red part

**degree of freedom for t test: -2 because means of two sample groups**

**df for CI: -1 because calculated for one group?**

### Mice

pair t test **(but n \>30?)**

H0: there is no difference

two tail test

assumption: normality. **no variance requirement since there are the same sample at different time**

only focus on the change, thus, 1 sample, df = 39

```{r}
## test the normality of the difference of the weight (after-before)
mice_dt = read.csv("class notes/mice_dt.csv")
mice_diff = with(mice_dt, af_dat - bf_dat)

with(mice_dt, t.test(bf_dat, af_dat, paired = T))
## the t-test can also be runned as a one-sample t-test and compared to 0

```

**with() function usage?**

## Type1 2 errors

focusing on the definitions

```{r}
# type 1 error (need to ensure this)
alpha = 0.05

# find quantile
qnorm(p = 0.95, mean = 0, sd = 0.4)

# type 2 error
beta = pnorm(0.66, 1, 0.4)
power = pnorm(0.66, 1, 0.4, lower.tail = F)

# power = 1-beta
```

## River

```{r}

river = read.csv("class notes/River_pH.csv")

rivA = river$pH[river$River_name == "A"]
rivB = river$pH[river$River_name == "B"]

# test the similarity of the sds
ratio = sd(rivA) / sd(rivB)
## or
var.test(rivA, rivB, alternative = "two.sided")
## conclusion: no significant difference, assumption holds

# t test
t.test(rivA, rivB, var.equal = T, alternative = "two.sided")
```

```{r}
t.test(rivA, rivB, var.equal = T, alternative = "g")

# how to calculate the CIs for the two sample tests?
```

```{r}
## assume A and B are paired tests (assume sd the same, only test normality)
# one sample (df = 9) 1 sided

t.test(rivB, rivA, paired = T, alternative = "l")

# or
t.test(river$pH ~ river$River_name, paired = "T", alternative = "greater")
```

```{r}
t.test(river$pH, mu = 7, alternative = "g")
```
