### Tree mass

```{r}
lmer(agb ~ ht + dbh + dist + (1|plot) + (1|species), data = data)
```

treat the species and plot as random effects since they are randomly chosen from a large population

visualization: right skewed of each variable (do not need to focus on the normality?? only concerns about the normality of the residuals, as shown in the plot later on)

can also see patterns of each two variables

evaluate the average effects of plots and species, and each of them varies a little than the average (reflected by the variances)

```{r}
# install.packages("performance")

```

reqq: looks bad, they are not linearly related

can try the log transformation of dept variable

the reqq better? because most of them is better, and only several outliers influence the shape?

use residual \< -1 to detect the records of outliers, and see are they make sense or not (too high tree; mismatch of height and width(DBH))

use (Plot \| species) to see the effect of their interaction

when to use this? can use (1\|plot) and this term together in a model?

why suddenly log transform a indpt variable?

what the work above is about: transformation of variables and remove outliers

compare model: the same dataset with different choices of variables

explain the meaning of the log-transformed model

log-transformed variables: the amount of change is described in **percent** rather than original unit

**How to see the proportion of variance explained by the model's main effect and random effects?**

MuMIn package R squared

Then, scaled the data

why do this? to better see the random effects, as shown in the plots on the second slide?

Prediction

```{r}
biomass C/N ~ as.factor(diversity) + as.factor(success) + mineralization + nitrification + denitrification + (1|plot) + (1|site) + as.factor(seasons) + as.factor(interested envir factors)
```

intro 1 page single-space academic paper format

如果对某些影响因素不感兴趣，就不要在模型里包括它们 e.g. other envir factors

除非它们是nature of the experiment design e.g. plot, site

temporal factors不一定是模型设计的一部分，因为有的实验并不关心temporal changes
