---
title: "3.8 linear model4"
author: "Jiahuan Li"
date: "2023-03-08"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## EPA gas standards

change the categorical variables only cause changes in the intercept (they do not influence the x)

about the log-transforms? (do not catch the key points) it is useful since it makes the relationship linear?

the log transform does not change the squared R

## Model selection

comparison limits:

1.  the models should based on the same dataset
2.  should not compare the log-transformed models to not transformed ones.

```{r}
kid <- read.csv("data/kidiq-1.csv")
model <- lm(kid.score ~ factor(mom.hs) + mom.iq + mom.age + factor(mom.work), kid)
# intercept: hs = 0, work = 1, iq = 0 (continuous variable)
summary(model)

step(model)
# additional to the step(), the process can also be done mannually. 
# Remove all the insignificant independent variables (only refer to the p-value, do not consider the squared-R). And run the new model (update the model step by step)
# keep hs and iq only
# the R-squared won't change a lot, meaning removing the vars do not matter
```
