---
title: 'ENV 710: Lab 4'
author: "Jiahuan Li"
date: "Feb 14, 2023"
output:
  pdf_document: default
  word_document: default
header-includes: \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=F, eval=T, message = FALSE, fig.pos = 'H')
require(ggplot2)
require(moments)
```

<!-- sample code chunks included for problems, add or remove any code chunks or change the settings -->

## Problem 1

### `myqqplot` function

```{r myqqplot, echo=TRUE}
#my qqplot function

myqqplot <- function(c){
  d <- sort(c)
  n <- length(d)
  
  p <- numeric(n)
  qZ <- numeric(n)
  
  for (i in 1:n){
    p[i] <- (i - 0.5)/n
    qZ[i] <- qnorm(p[i], mean = 0, sd = 1)
  }
  
  i1 <- 0.25 * n + 0.5
  i2 <- 0.75 * n + 0.5
  x1 <- qZ[i1]
  x2 <- qZ[i2]
  y1 <- d[i1]
  y2 <- d[i2]
  
  m <- (y2-y1)/(x2-x1)
  b <- y2 - m*x2
  
  
  fig <- plot (qZ, d, xlab = "Theoretical quantiles", ylab = "Sample values") + abline(b, m)
  return(fig)
}
```

### (a) `qqplot` generated using `myqqplot` function:

```{r problem1a, fig.align='center', fig.cap="\\label{fig:figs}Problem 1a", out.width='60%'}
# Problem 1 (a) 
AfrPlots <- read.csv("labs/lab4 Confidence Intervals/AfrPlots.csv")
df1 <- subset(AfrPlots, AfrPlots$CensusNo == 2)

myqqplot(df1$Dead)
```

### (b) `qqplot` generated using `qqnorm` and `qqline`:

```{r problem1b, fig.cap="\\label{fig:figs}Problem 1b", out.width='60%', fig.align='center'}
# Problem 1 (b) 

qqnorm(df1$Dead, las=1, main ="")
qqline(df1$Dead)

# my q-q plot look the same as that produced by qqnorm
```

Processing summary of my qqplot: at first we should understand the meaning of x and y axis. If we do not want to calculate Z scores of the original data to standardize the graph (making the y values also distributed in the interval around (-2,2)), the y values should just be the original observation values.

As for the calculation of x values, we should first find their linkages to y values, which is that they share the same percentile value. Thus, we should find the percentile of each observation by sorting and using the formula (i - 0.5)/n. Then, according to the corresponding percentile, we are able to calculate theoretical quantiles as the x values for the standardized normal distribution. And the -0.5 in the formula is applied to avoid the mismatch of 100% percentile which is infinity in the normal distribution.

\newpage

## Problem 2

### (a)

```{r problem2a, fig.cap="\\label{fig:figs}Density Histogram", out.width='70%', message=FALSE, fig.align='center'}
# Problem 2 (a) 
epa <- read.csv("labs/lab4 Confidence Intervals/epa2012.csv")
data <- epa$hwy_mpg

# visualization
ggplot(data.frame(data), aes(x=data)) + 
  geom_histogram(aes(y = after_stat(density)), 
  colour="white", fill = "blue", alpha = 0.7) +
  geom_density(size = 1) + xlab("Highway Gas Mileage") + 
  ylab("Density") + theme_bw()
```

```{r problem2b, fig.cap="\\label{fig:figs}Q-Q plot", out.width='60%', fig.align='center'}
# Problem 2 (b) 
ggplot(data.frame(data), aes(sample=data)) +
stat_qq() + stat_qq_line() +
ylab("Sample quantiles") + xlab("Theoretical quantiles") + theme_bw()
```

```{r problem2c, fig.cap="\\label{fig:figs}boxplot", out.width="60%", fig.align='center'}
# Problem 2 (c) 
ggplot(data.frame(data), aes(x = data, y = "")) + 
  stat_boxplot(geom = "errorbar", width = 0.15, color = 1) + 
  geom_boxplot(fill = "blue", alpha = 0.5, color = 1, outlier.colour = 2) + 
  xlab("Highway Gas Mileage") + 
  ylab("") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  theme_bw()
```

```{r problem2d}
# Problem 2 (d)

skewness <- function(y) {
  n <- length(y)
  skew <- 1/(n*sd(y)^3)*sum((y-mean(y))^3)
  return(skew)
}

kurtosis <- function(y) {
    n <- length(y)
    kurt <- 1 / (n * sd(y) ^ 4) * sum((y - mean(y, na.rm = TRUE)) ^ 4) - 3
    kurt
}

ans1 <- skewness(data)

ans2 <- kurtosis(data)

print(paste("The skewness is", ans1))
print(paste("The kurtosis is", ans2))
```

### (b)

```{r problem2alog, fig.align='center', fig.cap="\\label{fig:figs}Logarithmic Dencity Histogram",out.width='70%'}
# Problem 2 (a --- log) 
log_da <- log(data)

ggplot(data.frame(log_da), aes(x=log_da)) + 
  geom_histogram(aes(y = after_stat(density)), 
  colour="white", fill = "blue", alpha = 0.7) +
  geom_density(size = 1) + xlab("Highway Gas Mileage") + 
  ylab("Density") + theme_bw()
```

```{r problem2blog, fig.align='center', fig.cap="\\label{fig:figs}Logarithmic Q-Q Plot",out.width='60%'}
# Problem 2 (b --- log) 
ggplot(data.frame(log_da), aes(sample=log_da)) +
stat_qq() + stat_qq_line() +
ylab("Sample quantiles") + xlab("Theoretical quantiles") + theme_bw()
```

```{r problem2clog, fig.align='center', fig.cap="\\label{fig:figs}Logarithmic Boxplot",out.width='60%'}
# Problem 2 (c --- log) 
ggplot(data.frame(log_da), aes(x = log_da, y = "")) + 
  stat_boxplot(geom = "errorbar", width = 0.15, color = 1) + 
  geom_boxplot(fill = "blue", alpha = 0.5, color = 1, outlier.colour = 2) + 
  xlab("Highway Gas Mileage") + 
  ylab("") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  theme_bw()
```

```{r problem2dlog}
# Problem 2 (d --- log) 
ans3 <- skewness(log_da)

ans4 <- kurtosis(log_da)

print(paste("The skewness is", ans3))
print(paste("The kurtosis is", ans4))
```

### (c) Conclusion

The log-transformed data looks more normally distributed

```{r problem2conclusion}
# Problem 2 (conclusion) 

# Mean
mean <- mean(log_da)

# CIs
z_c <- qnorm(0.975, mean = 0, sd = 1)
low_x <- mean - z_c*sd(log_da)/sqrt(length(log_da))
up_x <- mean + z_c*sd(log_da)/sqrt(length(log_da))

print (paste("The mean is", mean))
print (paste("THe CIs is (", low_x, ",", up_x, ")"))
```

\newpage

```{r}
labs = knitr::all_labels()
labs = setdiff(labs, c("setup", "get-labels"))

```
