---
output: html_document
editor_options: 
  chunk_output_type: console
---

## for loops

```{r}
myLet = c('d','u','k','e')
for (i in 1:4) {
  for (j in 1: length(myLet)) {
    let = myLet[j]
    print(paste(i,let,sep = "-"))
  }
}
```

## Matrix

create a matrix

```{r}
## the number of data should be equal to ncol * nrow
myMatrix = matrix(data = rnorm(100, 0, 1), ncol = 5, nrow = 20)
class(myMatrix)

# length of this matrix is 100
length(myMatrix)
```

create an empty vector to hold the mean of each row

```{r}
# seems like the defult mode of vector() command is "logical"
meanV = vector(mode = 'numeric' ,length = nrow(myMatrix))

help(vector)

class(meanV)
typeof(meanV)
```

loop over rows to get and store the mean of each row

```{r}
for (i in 1:length(nrow(myMatrix))) {
  meanV[i] = mean(myMatrix[i,])
}
meanV
```

visualize each column (create five pictures in the plots)

```{r}
# pch: point style
for (i in 1:ncol(myMatrix)) {
  plot (myMatrix[,i], main = paste("Iteration number", i), pch = 15)
}
```

## Dataframe

create a dataframe

```{r}
fruits = c('apple', 'orange', 'banana', 'nectarine')
dishes = c('waffles', 'pancakes', 'french toast', 'salad')

df = data.frame(Fruit = fruits, Dish = dishes)

# length of df equals to column numbers
length(df)
```

print each rows: `df$(col_name)[i]`

```{r}
for (i in 1:length(fruits)) {
  print(paste("I love", df$Fruit[i], "on", df$Dish[i]))
}
```

rate fruits and save the result to df

```{r}
# nchar: the sum number of the string characters
rate = function(fruit){
  rating = nchar(fruit)
  rating
}

for (i in 1:nrow(df)){
  df$rate[i] = rate(df$Fruit[i])
}
```

won't plot since the fruit is not numeric value

```{r}
plot(df$Fruit, df$rate)
ggplot(df, aes(Fruit, rate), gem_col(),xlab('Fruits'),ylab('rating'))
```

## subset dataframe

```{r}
epa = read.csv ("labs/lab4 Confidence Intervals/epa2012.csv", sep = ",")
```

#### method 1 using index

```{r}
# and
idx1 = which(epa$mfr_name == "Audi" & epa$comb_mpg > 25)
subset1 = epa[idx1, ]

# or
idx2 = which(epa$mfr_name == "Audi" | epa$comb_mpg > 25)
subset2 = epa[idx2, ]
```

#### method 2 subset function

```{r}
subset3 = subset(epa, epa$mfr_name == "Audi")
```

## Central limit theory (CLT)

```{r}
# a series of 0 for 10000 times
umeans <- numeric(10000)

set.seed(1001)

## calculate the sample means for 10000 samples
for (i in 1:10000){
# Take the mean of a sample of 5 uniformly distributed numbers
# and store it in the vector, umeans
umeans[i] <- mean(runif(5, min = 0, max = 10))
}
```

visualization

```{r}
# Make a histogram of the 10,000 means
gg1 <- ggplot(data.frame(umeans), aes(x=umeans)) +
geom_histogram(colour = "white", fill = "blue") +
ylab("Frequency") + xlab("Sample means") +
theme_bw()

# Give it a gradient of colors for fun
gg2 <- ggplot(data.frame(umeans), aes(x=umeans, fill = ..count..)) +
geom_histogram(colour = "white") +
ylab("") + xlab("Sample means") +
scale_fill_gradient(low = "green", high = "blue") + theme_bw() +
theme(legend.position="none")

# alignment
## install.packages("ggpubr", repos = "https://cloud.r-project.org/", dependencies = TRUE)
ggpubr::ggarrange(gg1, gg2, ncol = 2, labels = "auto")
```

### To do 1

```{r}
test = seq(1,20)
sqrt = numeric(20)
for (i in 1:20) {
  sqrt[i] = sqrt(test[i])
}

sqrt
```

## Testing for Normality
